Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-10 journal paper/content/related_work.tex
a pass through related work and remaining color edits

\xhdr{Exploration and competition.}
\asmargincomment{This whole xhdr is new.}
\citet{bergemann1997market,bergemann2000experimentation,keller2003price} studied the interplay of exploration and competition for users when the competing firms experiment with \emph{prices}, rather than design alternatives. All three papers consider strategies that respond to competition, and analyze Markov Perfect Equilibria (MPE), whereas we focus on the adoption of better bandit algorithms. We discuss them in more detail below.

@@

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing \emph{more} information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms'.%
\footnote{The planner faces a two-armed bandit problem, rather than a pricing problem. This is because the planner can directly choose a firm for each agent (by choosing a very low price for this firm, and a very high price for the other).}
% firm, because payments cancel out in social welfare).}
Consequently, when one compares competition to the planner, one can compare the outcomes but not the algorithms.

@@

%\gaedit{\citet{bergemann1997market,bergemann2000experimentation} studied a model where there is an incumbent with known product quality and an entrant with unknown, but exogenously determined, quality. Similar to \citet{keller2003price}, the firms compete on prices and they characterize the MPE of dynamic competition between the two firms. In their model, consumer purchases of the entrant's product result in an informational externality that allows both the firms and consumers to gather information on the entrant's product quality. However, we focus on a model where firms' product quality is endogenous and the focus of exploration in a context where the price mechanism is absent.}

\vspace{2mm}

\asmargincomment{It serves us to discuss this work here, IMO, and we have an excuse.}
In the line of work on \emph{strategic experimentation} (starting from \citet{Bolton-econometrica99,Keller-econometrica05}, see \citet{Horner-survey16} for a survey), agents explore and learn over time in a shared environment. Thus, we have exploration algorithms which interact with each other strategically, \eg each agent would prefer to free-ride on the exploration done by others. However, the agents do not compete with each other in any meaningful sense.

Several papers study competition between two principals who run algorithms but do not interact, directly or indirectly, until the very end of the game. \cite{Ufuk-jeea15} consider a ``research competition" between two firms racing towards a big discovery. 
\asmargincomment{This way, this paper seems not that relevant, doesn't it?}
Each firm deploys a bandit algorithm with two arms, corresponding to safe and risky lines of research. The firms do not interact untl one of them makes the discovery and wins the game. In the ``dueling algorithms" framework of \citet{DuelingAlgs-stoc11}, each principals runs an algorithm for the same problem. All inputs are observable at once, and principals' payoffs are binary (win/lose). \citet{ben2017best, ben2019regression} study competition between ``offline" machine learning algorithms. In comparison, we study a ``product competition" in which the two firms interact continuously (via the customers' choices), accrue rewards incrementally, and compete for individual customers.

\asmargincomment{This happens to be about exploration \& competition ...}
A long line of work from electrical engineering and computer science, starting from
\citet{MultiPlayerMAB-Poor08,MultiPlayerMAB-Liu10,MultiPlayerMAB-Anima11}, targets an application to \emph{cognitive radios}, where multiple radios transmit simultaneously in a shared medium and compete for bandwidth. Each radio chooses channels over time using a multi-armed bandit algorithm. This work studies a repeated game between bandit algorithms, and focuses on designing algorithms which work well in this game, under various assumptions on communication, synchronization and collisions.

@@

\asedit{\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". The role of data as a barrier to entry in online markets has been noted in \cite{de2020data}; we find a similar effect with ``data advantage". \citet{kerin1992first} overview the other channels through which first-mover advantage can lead to a competitive advantage.}

@@

\xhdr{Relaxed rationality.}
Relaxed versions of rationality similar to ours are found in several notable lines of work. For example, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. The \SoftMaxRandom model is closely related to the literature on \emph{product differentiation}, starting from \cite{Hotelling-29}, see \cite{Perloff-Salop-85} for a notable later paper. There is a large literature on non-existence of equilibria due to small deviations   (which is related to the corresponding result for \HardMaxRandom), starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. Notable recent papers \citep{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom.
\asedit{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

>>>>

\xhdr{Exploration and competition.}
\citet{bergemann1997market,bergemann2000experimentation,keller2003price} studied the interplay of exploration and competition for users when the competing firms experiment with \emph{prices}, rather than design alternatives. All three papers consider strategies that respond to competition, and analyze Markov Perfect Equilibria (MPE), whereas we focus on the adoption of better bandit algorithms. We discuss them in more detail below.

@@

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, \gaedit{exogenously determined}, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. \gadelete{Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low).} \gaedit{Third}, the social planner's exploration problem is very different from the firms' \gaedit{so that when one compares competition to the planner, one can only compare outcomes, not the algorithms.}\footnote{The planner faces a two-armed bandit problem, rather than a pricing problem \gaedit{because the planner can directly choose a firm for each agent.}}
% firm, because payments cancel out in social welfare).}

@@

%\gaedit{\citet{bergemann1997market,bergemann2000experimentation} studied a model where there is an incumbent with known product quality and an entrant with unknown, but exogenously determined, quality. Similar to \citet{keller2003price}, the firms compete on prices and they characterize the MPE of dynamic competition between the two firms. In their model, consumer purchases of the entrant's product result in an informational externality that allows both the firms and consumers to gather information on the entrant's product quality. However, we focus on a model where firms' product quality is endogenous and the focus of exploration in a context where the price mechanism is absent.}
In the line of work on \emph{strategic experimentation} (starting from \citet{Bolton-econometrica99,Keller-econometrica05}, see \citet{Horner-survey16} for a survey), agents explore and learn over time in a shared environment. Thus, we have exploration algorithms which interact with each other strategically, \eg each agent would prefer to free-ride on the exploration done by others. However, the agents do not compete with each other in any meaningful sense.

Several papers study competition between two principals who run algorithms but do not interact, directly or indirectly, until the very end of the game. \cite{Ufuk-jeea15} consider a ``research competition" between two firms racing towards a big discovery. 
Each firm deploys a bandit algorithm with two arms, corresponding to safe and risky lines of research. The firms do not interact until one of them makes the discovery and wins the game. In the ``dueling algorithms" framework of \citet{DuelingAlgs-stoc11}, each principals runs an algorithm for the same problem. All inputs are observable at once, and principals' payoffs are binary (win/lose). \citet{ben2017best, ben2019regression} study competition between ``offline" machine learning algorithms. In comparison, we study a ``product competition" in which the two firms interact continuously (via the customers' choices), accrue rewards incrementally, and compete for individual customers.

A long line of work from electrical engineering and computer science, starting from
\citet{MultiPlayerMAB-Poor08,MultiPlayerMAB-Liu10,MultiPlayerMAB-Anima11}, \gaedit{focuses on competition for resources, not competition for consumers. Specifically, this literature targets an application} to \emph{cognitive radios}, where multiple radios transmit simultaneously in a shared medium and compete for bandwidth. Each radio chooses channels over time using a multi-armed bandit algorithm. This work studies a repeated game between bandit algorithms, and focuses on designing algorithms which work well in this game, under various assumptions on communication, synchronization and collisions.

@@

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". The role of data as a barrier to entry in online markets has been noted in \cite{de2020data}; we find a similar effect with ``data advantage". \citet{kerin1992first} overview the other channels through which first-mover advantage can lead to a competitive advantage.

@@

\xhdr{Relaxed rationality.}
Relaxed versions of rationality similar to ours are found in several notable lines of work. For example, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. The \SoftMaxRandom model is closely related to the literature on \emph{product differentiation}, starting from \cite{Hotelling-29}, see \cite{Perloff-Salop-85} for a notable later paper. There is a large literature on non-existence of equilibria due to small deviations   (which is related to the corresponding result for \HardMaxRandom), starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. Notable recent papers \citep{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom.
While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.

### 2020-09-10 journal paper/content/sec-intro.tex
a pass through related work and remaining color edits

Learning from interactions with users is ubiquitous in modern customer-facing platforms, from product recommendations to web search to content selection to fine-tuning user interfaces. Many platforms purposefully implement \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a standard and well-studied framework for exploration and making decisions under uncertainty. \asedit{This trend has been stimulated by two factors: almost-zero cost of deploying iterations of a product (provided an initial infrastructure investment), and the fact that many online platforms primarily compete on product quality, rather than price.
(\eg because they are supported by ads or cheap subscriptions).}
%\citep{Gittins-book11,Bubeck-survey12,slivkins-MABbook,LS19bandit-book}.

@@

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
We interpret innovation as the adoption of better algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}
%, even if the algorithms themselves are well-known in the research literature
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.%
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
\asedit{The barriers for innovation arise entirely from the reputational consequences of exploration in competition, whereas the inverted-U relationships in prior work focus on the tradeoff between the benefits of the technology and its R\&D costs.}

@@

\xhdr{Significance.}
Our results have a dual purpose: shed light on real-world implications of some typical scenarios, and investigate the space of models for describing the real world. As an example for the latter: while the \HardMax model with simultaneous entry is arguably the most natural model to study \emph{a priori}, our results elucidate the need for more refined models with ``free exploration" (\eg via random agents or early entry). \asedit{On a technical level, we connect a literature on regret-minimizing bandits in computer science and that on competition in economics.}

>>>>

Learning from interactions with users is ubiquitous in modern customer-facing platforms, from product recommendations to web search to content selection to fine-tuning user interfaces. Many platforms purposefully implement \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. Online platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a standard and well-studied framework for exploration and making decisions under uncertainty. This trend has been stimulated by two factors: almost-zero cost of deploying iterations of a product (provided an initial infrastructure investment), and the fact that many online platforms primarily compete on product quality, rather than price.
(\eg because they are supported by ads or cheap subscriptions).
%\citep{Gittins-book11,Bubeck-survey12,slivkins-MABbook,LS19bandit-book}.

@@

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better. %\citep[\eg][]{aghion2005competition,Vives-08}.
We interpret innovation as the adoption of better algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}
%, even if the algorithms themselves are well-known in the research literature
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.%
%\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
The barriers for innovation arise entirely from the reputational consequences of exploration in competition, whereas the inverted-U relationships in prior work focus on the tradeoff between the benefits of the technology and its R\&D costs.

@@

\xhdr{Significance.}
Our results have a dual purpose: shed light on real-world implications of some typical scenarios, and investigate the space of models for describing the real world. As an example for the latter: while the \HardMax model with simultaneous entry is arguably the most natural model to study \emph{a priori}, our results elucidate the need for more refined models with ``free exploration" (\eg via random agents or early entry). On a technical level, we connect a literature on regret-minimizing bandits in computer science and that on competition in economics.

### 2020-09-10 journal paper/content/sec-model.tex
remove color edits and fix formatting of tables in the appendix
a pass through related work and remaining color edits

\xhdr{Extensions.} \asedit{The \TheoryModel admits several extensions, detailed in Section~\ref{sec:theory-extensions}. First, all/most results extend to arbitrary reward distributions, allow reward-dependent utility, and carry over to a more general version of multi-armed bandits. Second, agents could have beliefs on $(\alg[1],\alg[2],\priorMu,\respF)$ that need not be correct; then, all results carry over with respect to these beliefs. 
%Third, all results carry over \asmargincomment{check}  if each agent $t$ has a prior on her arrival time $t$, rather than know it exactly.
Third, we can handle a limited amount of non-stationarity in $\respF$. \asmargincomment{check} Finally, the main result on \HardMax extends to time-discounted utilities.
}

@@

\asedit{Reflecting reality, we posit that principals' exploration strategies are ``non-strategic" in nature. Even though the principals play a multi-step game, they do not react to each other's moves or to the agents' strategic choices. This is how industry approaches exploration algorithms, and for several good reasons. "Non-strategic" exploration is well-studied in machine learning, and yet it remains a very complex subject in research (as evidenced, \eg by the huge amount of activity therein). Even the seemingly simple algorithms are not straightforward to deploy in practice, and require a substantial investment in infrastructure \cite[\eg see the discussions in][]{DS-arxiv}. Responding to the competition represents another layer of complexity which has not been previously studied in this context, to the best of our knowledge, let alone made even remotely practical. Besides, the competitor's exploration strategy is typically not public, and understanding its exploration behavior via observations appears challenging even as a  research problem.}%
\footnote{\asedit{Principals could potentially react to the market share or (the difference in) reputation scores. However, baking these signals into one's exploration strategy runs the risk of over-interpreting our competition model, as they may change for exogenous reasons. Alternatively, one could use such signals, as well as the intuitions coming from this paper, to guide the platform's decisions regarding exploration.}}

@@

\asedit{On the machine learning side, our model captures big, qualitative differences between bandit algorithms, building on the well-established intuition in the literature. Comparisons between bandit algorithms are generally somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. In particular, ``better" algorithms are better in the long run, but could be worse initially. We focus on a standard model of stochastic bandits. In Appendix~\ref{app:bg}, we present sufficient background on this model, accessible to non-specialists. However, we are less interested in state-of-art algorithms for realistic applications.}

\asedit{Bandit rewards are not discounted with time, reflecting the fact that non-discounted, regret-minimizing formulations has been prevalent in the bandits literature over the past two decades \citep{slivkins-MABbook,LS19bandit-book}, and better correspond to practical deployments \citep[\eg][]{DS-arxiv}. Also, the distinctions between better and worse bandit algorithms are not as well-understood under time-discounting.}

>>>>

\xhdr{Extensions.} \asedit{The \TheoryModel admits several extensions, detailed in Section~\ref{sec:theory-extensions}. First, all/most results extend to arbitrary reward distributions, allow reward-dependent utility, and carry over to a more general version of multi-armed bandits. Second, agents could have beliefs on $(\alg[1],\alg[2],\priorMu,\respF)$ that need not be correct; then, all results carry over with respect to these beliefs. 
%Third, all results carry over \asmargincomment{check}  if each agent $t$ has a prior on her arrival time $t$, rather than know it exactly.
Third, we can handle a limited amount of non-stationarity in $\respF$ for the \HardMaxRandom and \SoftMaxRandom decision rules. \asmargincomment{check} Finally, the main result on \HardMax extends to time-discounted utilities.
}

@@

Reflecting reality, we posit that principals' exploration strategies are ``non-strategic" in nature. Even though the principals play a multi-step game, they do not react to each other's moves or to the agents' strategic choices. This is how industry approaches exploration algorithms, and for several good reasons. "Non-strategic" exploration is well-studied in machine learning, and yet it remains a very complex subject in research (as evidenced, \eg by the huge amount of activity therein). Even the seemingly simple algorithms are not straightforward to deploy in practice, and require a substantial investment in infrastructure \cite[\eg see the discussions in][]{DS-arxiv}. Responding to the competition represents another layer of complexity which has not been previously studied in this context, to the best of our knowledge, let alone made even remotely practical. Besides, the competitor's exploration strategy is typically not public, and understanding its exploration behavior via observations appears challenging even as a  research problem.%
\footnote{Principals could potentially react to the market share or (the difference in) reputation scores. However, baking these signals into one's exploration strategy runs the risk of over-interpreting our competition model, as they may change for exogenous reasons. Alternatively, one could use such signals, as well as the intuitions coming from this paper, to guide the platform's decisions regarding exploration.}

@@

On the machine learning side, our model captures big, qualitative differences between bandit algorithms, building on the well-established intuition in the literature. Comparisons between bandit algorithms are generally somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. In particular, ``better" algorithms are better in the long run, but could be worse initially. We focus on a standard model of stochastic bandits. In Appendix~\ref{app:bg}, we present sufficient background on this model, accessible to non-specialists. However, we are less interested in state-of-art algorithms for realistic applications.

Bandit rewards are not discounted with time, reflecting the fact that non-discounted, regret-minimizing formulations has been prevalent in the bandits literature over the past two decades \citep{slivkins-MABbook,LS19bandit-book}, and better correspond to practical deployments \citep[\eg][]{DS-arxiv}. Also, the distinctions between better and worse bandit algorithms are not as well-understood under time-discounting.

### 2020-09-10 journal paper/content/sec-theory.tex
a pass through related work and remaining color edits

>>>>

\xhdr{Beliefs.} Rather than knowing the principals' algorithms $(\alg[1],\alg[2])$, the Bayesian prior $\priorMu$, and the response function $\respF$, agents could have beliefs on these objects that need not be correct. If agents have common ``point beliefs", respectively, algorithms $(\alg[1]',\alg[2]')$ prior $\priorMu'$ and response function $\respF'$,  then all our results carry over with respect to these beliefs.

\xhdr{Limited non-stationarity in $\respF$.} 
Different agents can have different response functions, in the following sense. For \HardMaxRandom, our results carry over if each agent $t$ has a \HardMaxRandom response function $\respF$ with parameter $\eps_t\geq \eps_0$. For \SoftMaxRandom, different agents can have different response functions that satisfy Definition~\ref{def:SoftMax} (with the same parameters).
